### 2020-08-03 submission/manuscript.tex
Incorporate Pat's and my own edits

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{}
    \pretitle{\vspace{\droptitle}}
  \posttitle{}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

@@

@@

\section{\texorpdfstring{The initial gut microbiota and response to
antibiotic perturbation influence \emph{Clostridioides difficile}
colonization in
mice}{The initial gut microbiota and response to antibiotic perturbation influence Clostridioides difficile colonization in mice}}\label{the-initial-gut-microbiota-and-response-to-antibiotic-perturbation-influence-clostridioides-difficile-colonization-in-mice}

@@

@@

\linenumbers

\subsection{Abstract}\label{abstract}

The gut microbiota has a key role in determining susceptibility to
\emph{Clostridioides difficile} infections (CDIs). However, much of the
mechanistic work examining CDIs in mouse models use animals obtained
from a single university colony or vendor. We treated mice from 6
different sources (2 University of Michigan colonies and 4 vendors) with
a single clindamycin dose, followed by a \emph{C. difficile} challenge 1
day later and then measured \emph{C. difficile} colonization levels
through 9 days post-infection. The microbiota was profiled via 16S rRNA
gene sequencing to examine the variation across sources and alterations
due to clindamycin treatment and \emph{C. difficile} challenge. While
all sources of mice were colonized 1-day post-infection, variation
emerged from days 3-7 post-infection with animals from some sources
colonized with \emph{C. difficile} for longer and at higher levels. We
identified bacteria that varied in relative abundance across sources and
throughout the experiment. Some bacteria were consistently impacted by
clindamycin treatment in all sources of mice including
\emph{Lachnospiraceae}, \emph{Ruminococcaceae}, and
\emph{Enterobacteriaceae}. To identify bacteria that were most important
to colonization regardless of the source, we created logistic regression
models that successfully classified mice based on whether they cleared
\emph{C. difficile} by 7 days post-infection using baseline,
post-clindamycin, and post-infection community composition data. With
these models, we identified 4 bacteria that varied across sources
(\emph{Bacteroides}), were altered by clindamycin
(\emph{Porphyromonadaceae}), or both (\emph{Enterobacteriaceae} and
\emph{Enterococcus}). Microbiota variation across sources better
emulates human interindividual variation and can help identify bacterial
drivers of phenotypic variation in the context of CDIs.

\subsection{Importance}\label{importance}

\emph{Clostridioides difficile} is a leading nosocomial infection.
Although perturbation to the gut microbiota has been established as a
key risk factor, there is variation in who becomes asymptomatically
colonized, develops an infection, or has an infection with adverse
outcomes. \emph{C. difficile} infection (CDI) mouse models are widely
used to answer a variety of \emph{C. difficile} pathogenesis questions.
However, the inter-individual variation between mice is less than what
is observed in humans, particularly if just one source of mice is used.
In this study, we administered clindamycin to mice from 6 different
breeding colonies and challenged them with \emph{C. difficile}.
Interestingly, only a subset of the bacteria that vary across sources
were associated with how long \emph{C. difficile} was able to colonize.
Future studies examining the interplay between the microbiota and
\emph{C. difficile} should consider using mice from multiple sources to
better reflect human interindividual variation.

@@

\subsection{Introduction}\label{introduction}

Antibiotics are a common risk factor for \emph{Clostridioides difficile}
infections (CDIs), but there is variation in who goes on to develop
severe or recurrent CDIs after exposure (1, 2). Additionally,
asymptomatic colonization, where \emph{C. difficile} is detectable, but
symptoms are absent has been documented in infants and adults (3, 4).
The intestinal microbiome has been implicated in asymptomatic
colonization (5, 6), susceptibility to CDIs (7), and adverse CDI
outcomes (9--12).

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

@@

\author{}
\date{\vspace{-2.5em}}

@@

\hypertarget{the-initial-gut-microbiota-and-response-to-antibiotic-perturbation-influence-clostridioides-difficile-colonization-in-mice}{%
\section{\texorpdfstring{The initial gut microbiota and response to
antibiotic perturbation influence \emph{Clostridioides difficile}
colonization in
mice}{The initial gut microbiota and response to antibiotic perturbation influence Clostridioides difficile colonization in mice}}\label{the-initial-gut-microbiota-and-response-to-antibiotic-perturbation-influence-clostridioides-difficile-colonization-in-mice}}

@@

\newpage
\linenumbers

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

The gut microbiota has a key role in determining susceptibility to
\emph{Clostridioides difficile} infections (CDIs). However, much of the
mechanistic work examining CDIs in mouse models use animals obtained
from a single university colony or vendor. We treated mice from 6
different sources (2 University of Michigan colonies and 4 vendors) with
a single clindamycin dose, followed by a \emph{C. difficile} challenge 1
day later and then measured \emph{C. difficile} colonization levels
through 9 days post-infection. The microbiota were profiled via 16S rRNA
gene sequencing to examine the variation across sources and alterations
due to clindamycin treatment and \emph{C. difficile} challenge. While
all sources of mice were colonized 1-day post-infection, variation
emerged from days 3-7 post-infection with animals from some sources
colonized with \emph{C. difficile} for longer and at higher levels. We
identified bacteria that varied in relative abundance across sources and
throughout the experiment. Some bacteria were consistently impacted by
clindamycin treatment in all sources of mice including
\emph{Lachnospiraceae}, \emph{Ruminococcaceae}, and
\emph{Enterobacteriaceae}. To identify bacteria that were most important
to colonization regardless of the source, we created logistic regression
models that successfully classified mice based on whether they cleared
\emph{C. difficile} by 7 days post-infection using baseline,
post-clindamycin, and 1-day post-infection community composition data.
With these models, we identified 4 bacteria that varied across sources
(\emph{Bacteroides}), were altered by clindamycin
(\emph{Porphyromonadaceae}), or both (\emph{Enterobacteriaceae} and
\emph{Enterococcus}). Microbiota variation across sources better
emulates human inter-individual variation and can help identify
bacterial drivers of phenotypic variation in the context of CDIs.

\hypertarget{importance}{%
\subsection{Importance}\label{importance}}

\emph{Clostridioides difficile} is a leading nosocomial infection.
Although perturbation to the gut microbiota is an established risk,
there is variation in who becomes asymptomatically colonized, develops
an infection, or has an infection with adverse outcomes. \emph{C.
difficile} infection (CDI) mouse models are widely used to answer a
variety of \emph{C. difficile} pathogenesis questions. However, the
inter-individual variation between mice from the same breeding facility
is less than what is observed in humans. Therefore, we challenged mice
from 6 different breeding colonies with \emph{C. difficile}. We found
that the starting microbial community structures and \emph{C. difficile}
persistence varied by the source of mice. Interestingly, a subset of the
bacteria that varied across sources were associated with how long
\emph{C. difficile} was able to colonize. By increasing the
inter-individual diversity of the starting communities, we were able to
better model human diversity. This provided a more nuanced perspective
of \emph{C. difficile} pathogenesis.

@@

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

Antibiotics are a common risk factor for \emph{Clostridioides difficile}
infections (CDIs) due to there effect on the intestinal microbiota, but
there is variation in who goes on to develop severe or recurrent CDIs
after exposure (1, 2). Additionally, asymptomatic colonization, where
\emph{C. difficile} is detectable, but symptoms are absent has been
documented in infants and adults (3, 4). The intestinal microbiota has
been implicated in asymptomatic colonization (5, 6), susceptibility to
CDIs (7), and adverse CDI outcomes (9--12). However, it is not clear how
much inter-individual microbiota variation contributes to the range of
outcomes observed after \emph{C. difficile} exposure relative to other
risk factors.

Mouse models of CDIs have been a great tool for understanding \emph{C.
difficile} pathogenesis (13). The number of CDI mouse model studies has
grown substantially since Chen et al.~published their C57BL/6 model in
2008, which disrupted the gut microbiota with antibiotics to enable
\emph{C. difficile} colonization and symptoms such as diarrhea and
weight loss (14). CDI mouse models have been used to examine
translationally relevant questions regarding \emph{C. difficile},
including the role of the microbiota and efficacy of potential
therapeutics for treating CDIs (15). However, variation in the
microbiota between mice from the same breeding colony is much less than
the inter-individual variation observed between humans (16, 17).
Studying CDIs in mice with a homogenous microbiota is likely to
overstate the importance of individual mechanisms. Using mice that have
a more heterogenous microbiota would allow researchers to identify and
validate more generalizable mechanisms responsible for CDI.

In the past, our group has attempted to introduce more variation into
the mouse microbiota by using a variety of antibiotic treatments
(18--21). An alternative approach to maximize microbiota variation is to
use mice from multiple sources (22, 23). The differences between the
microbiota of mice from vendors have been well documented and shown to
influence susceptibility to a variety of diseases (24, 25), including
enteric infections (22, 23, 26--30). Different research groups have also
observed variation in CDI outcomes despite using similar murine models
(13, 18, 21, 31--33). Here we examined how variations in the baseline
microbiota and responses to clindamycin treatment in C57BL/6 mice from
six different sources influenced susceptibility to \emph{C. difficile}
colonization and the time needed to clear the infection.

\hypertarget{results}{%
\subsection{Results}\label{results}}

\textbf{The variation in the microbiota is high between mice from
different sources.} We obtained C57BL/6 mice from 6 different sources:
two colonies from the University of Michigan that were split from each
other in 2010 (the Young and Schloss lab colonies) and four commercial
sources: the Jackson Laboratory, Charles River Laboratories, Taconic
Biosciences, and Envigo (which was formerly Harlan). These 4 vendors
were chosen because they are commonly used for murine CDI studies (26,
34--40). Two experiments were conducted, approximately 3 months apart.

We sequenced the 16S rRNA gene from fecal samples collected from these
mice after they acclimated to the University of Michigan animal housing
environment. We first examined the alpha diversity across the 6 sources
of mice. There was a significant difference in the richness (i.e.~number
of observed operational taxonomic units (OTUs)), but not Shannon
diversity index (\emph{P}\textsubscript{FDR} = 0.03 and
\emph{P}\textsubscript{FDR} = 0.052, respectively) across the sources of
mice (Fig. 1A-B and Tables S1-2). Next, we compared the community
structure of mice (Fig. 1C). Interactions between the source and cage
effects explained most of the observed variation between fecal
communities (PERMANOVA combined R\textsuperscript{2} = 0.90, \emph{P}
\textless{} 0.001, Fig. 1C and Table S3). Mice that are co-housed tend
to have similar gut microbiotas due to coprophagy (41). Since mice
within the same source were housed together, it is not surprising that
the cage effect also contributed to the observed microbiota variation.
There were some differences between the 2 experiments we conducted, as
the experiment and cage effects significantly explained the observed
community variation for the Schloss and Young lab mouse colonies (Fig.
S2A-B and Table S4). However, most of the vendors also clustered by
experiment (Fig. S2C-D, F), suggesting there was some community
variation between the 2 experiments within each source, particularly for
Schloss, Young, and Envigo mice (Fig. S2G-H). After finding differences
at the community level, we next identified the bacteria that varied
between sources of mice. There were 268 OTUs with relative abundances
that were significantly different between the sources (Fig. 1D and Table
S5). Though we saw differences between experiments at the community
level, there were no OTUs that were significantly different between
experiments within Schloss, Young, and Envigo mice (Data not shown). By
using mice from six sources we were able to increase the number of
microbiota communities to treat with clindamycin and challenge with
\emph{C. difficile}.

\textbf{Clindamycin treatment renders all mice susceptible to \emph{C.
difficile} 630 colonization, but clearance time varies across sources.}
Clindamycin is frequently implicated with human CDIs (42) and was part
of the antibiotic treatment for the frequently cited 2008 CDI mouse
model (14). We have previously demonstrated mice are rendered
susceptible to \emph{C. difficile}, but cleared the pathogen within 9
days when treated with clindamycin alone (21, 43). All mice were treated
with 10 mg/kg clindamycin via intraperitoneal injection and one day
later challenged with 10\textsuperscript{3} \emph{C. difficile} 630
spores (Fig. 2A). The day after infection, \emph{C. difficile} was
detectable in all mice at a similar level (median CFU range:
2.2e+07-1.3e+08; \emph{P}\textsubscript{FDR} = 0.15), indicating
clindamycin rendered all mice susceptible regardless of source (Fig.
2B). However, between 3 and 7 days post-infection, we observed variation
in \emph{C. difficile} levels across sources of mice (all
\emph{P}\textsubscript{FDR} \(\le\) 0.019; Fig. 2B and Table S6). This
suggested the mouse source was associated with \emph{C. difficile}
clearance. While the colonization dynamics were similar between the two
experiments, the Schloss mice took longer to clear in the 1st experiment
and the Envigo mice took longer to clear in the 2nd experiment (Fig.
S2A-B). The change in the mice's weight significantly varied across
sources of mice with the most weight lost two days post-infection (Fig.
2C and Table S7). There was also one Jackson and one Envigo mouse that
died between 1- and 3-days post-infection during the second experiment.
Mice obtained from Jackson, Taconic, and Envigo tended to lose more
weight, have higher \emph{C. difficile} CFU levels and take longer to
clear the infection compared to the other sources of mice (although
there was variation between experiments with Schloss and Envigo mice).
This was particularly evident 7 days post-infection (Fig. 2B-C, Fig.
S2C-D), when 57\% of the mice were still colonized with \emph{C.
difficile} (Fig. S2E). By 9 days post-infection the majority of the mice
from all sources had cleared \emph{C. difficile} (Fig. 2C) with the
exception of 1 Taconic mouse from the first experiment and 2 Envigo mice
from the second experiment. Thus, clindamycin rendered all mice
susceptible to \emph{C. difficile} 630 colonization, regardless of
source, but there was significant variation in disease phenotype across
the sources of mice.

\textbf{Clindamycin treatment alters bacteria in all sources, but a
subset of bacterial differences across sources persists.} Given the
variation in mouse microbiotas that we observed across breeding
colonies, we hypothesized that variation in \emph{C. difficile}
clearance would be explained by microbiota variation across the 6
sources of mice. As expected, clindamycin treatment decreased the
richness and Shannon diversity across all sources of mice (Fig. 3A-B).
Interestingly, significant differences in diversity metrics between
sources (\emph{P}\textsubscript{FDR} \textless{} 0.05) emerged after
clindamycin treatment, with Charles River mice having higher richness
and Shannon diversity than most of the other sources (Fig 3A-B and
Tables S1-2). The clindamycin treatment decreased the variation in
community structures between sources of mice. Source and cage effects
explained almost all of the observed variation between communities
(combined R\textsuperscript{2} = 0.99, \emph{P} \textless{} 0.001; Fig.
3C and Table S3). However, there were only 18 OTUs (Fig. 3D and Table
S8) with relative abundances that significantly varied between sources.
Next we identified the bacteria that shifted after clindamycin
treatment, regardless of source by analyzing paired fecal samples from
mice that were collected at baseline and after clindamycin treatment. We
identified 153 OTUs that were altered after clindamycin treatment in
most mice (Fig. 3E and Table S9). When we compared the list of
significant clindamycin impacted bacteria with the bacteria that varied
between sources post-clindamycin, we found 4 OTUs
(\emph{Enterobacteriaceae (OTU 1)}, \emph{Lachnospiraceae} (OTU 130),
\emph{Lactobacillus} (OTU 6), \emph{Enterococcus} (OTU 23)) overlapped
(Fig. 3D-E and Tables S8-9). Importantly, some of the OTUs that varied
between sources also shifted with clindamycin treatment. For example,
\emph{Proteus} increased after clindamycin treatment (Fig. 3D), but only
in Taconic mice. \emph{Enterococcus} was primarily found only in mice
purchased from commercial vendors and also increased after clindamycin
treatment (Fig. 3D). These findings demonstrate that clindamycin had a
consistent impact on the fecal bacterial communities of mice from all
sources and only a subset of the OTUs continued to vary between sources.

\textbf{Microbiota variation between sources is maintained after
\emph{C. difficile} challenge.} One day post-infection, significant
differences in diversity metrics remained across sources
(\emph{P}\textsubscript{FDR} \textless{} 0.05, Fig 4A-B and Tables
S1-2). Although the Charles River mice had more diverse microbiotas and
were also able to clear \emph{C. difficile} faster than some of the
other sources. Microbiota diversity did not explain the observed
variation in \emph{C. difficile} colonization across sources considering
the Young and Schloss mice had the lowest diversity after clindamycin
treatment and were able to clear \emph{C. difficile} earlier than
Jackson, Taconic and Envigo mice. Source and the interactions between
source and cage effects continued to explain most of the observed
community variation (combined R\textsuperscript{2} = 0.88; \emph{P}
\textless{} 0.001; Fig. 4D and Table S3). One day after \emph{C.
difficile} challenge, there were 44 OTUs (Fig. 4D and Table S10) with
significantly different relative abundances across sources.

Throughout the experiment the source of mice continued to be the
dominant factor that explained the observed variation across fecal
communities (PERMANOVA R\textsuperscript{2} = 0.35, \emph{P} \textless{}
0.001) followed by interactions between cage effects and the day of the
experiment (Movie S1 and Table S11). Mice fecal samples from the same
source of mice continued to cluster closely to each other throughout the
experiment. By 7 days post-infection, when approximately 43\% mice had
cleared \emph{C. difficile}, most of the mice still had not recovered to
their baseline community structure (Fig. 4E). Distance from the baseline
community between sources did not explain the variation in \emph{C.
difficile} clearance since Schloss and Young mice cleared \emph{C.
difficile} faster, but their communities were a greater distance from
baseline 7 days post-infection. In summary, mouse bacterial communities
varied significantly between sources throughout the course of the
experiment and a consistent subset of bacterial taxa remained different
between sources regardless of clindamycin and \emph{C. difficile}
challenge.

\textbf{Baseline, post-clindamycin, and post-infection community data
can predict mice that will clear \emph{C. difficile} by 7 days
post-infection.} After identifying taxa that varied between sources,
changed after clindamycin treatment, or both, we next wanted to
determine which taxa were influencing the variation in \emph{C.
difficile} colonization at day 7 (Fig. 2B, Fig. S2C). We trained three
L2-regularized logistic regression models with input bacterial community
data from the baseline (day = -1), post-clindamycin (day = 0), and
post-infection (day = 1) timepoints of the experiment to predict
\emph{C. difficile} colonization status on day 7 (Fig. S3A-B). All
models were better at predicting \emph{C. difficile} colonization status
on day 7 than random chance (all \emph{P} \textless{} 0.001, Table S12).
The model based on the post-clindamycin (day 0) community OTU data
performed significantly better than the other models with an area under
the receiving operator characteristic curve (AUROC) of 0.78
(\emph{P}\textsubscript{FDR} \textless{} 0.001 for pairwise comparisons,
Table S13). Thus, we were able to use bacterial relative abundance data
from the time of \emph{C. difficile} challenge to differentiate mice
that had cleared \emph{C. difficile} before day 7 from the mice still
colonized with \emph{C. difficile} at that timepoint. This result
suggests the bacterial community's response to clindamycin treatment had
the greatest influence on subsequent \emph{C. difficile} colonization
dynamics.

To examine the bacteria that were driving each model's performance, we
selected the 20 OTUs that had the highest absolute feature weights in
each of the 3 models (Table S14). First, we looked at OTUs from the
model with the best performance, which was based on the post-clindamycin
treatment (day 0) bacterial community data. Out of the 10 highest ranked
OTUs, 7 OTUs (\emph{Bacteroides}, \emph{Escherichia/Shigella}, 2
\emph{Lachnospiraceae}, \emph{Lactobacillus}, \emph{Porphyromonadaceae},
and \emph{Ruminococcaceae}) were associated with \emph{C. difficile}
colonization 7 days post-infection, while 3 OTUs
(\emph{Enterobacteriaceae}, \emph{Lachnospiraceae},
\emph{Porphyromonadaceae}) were associated with clearance (Fig. 5A).
Next, we examined whether any of the top 20 ranked OTUs from the
baseline (day 0) model were also important in the other 2 classification
models based on baseline (day -1) and 1 day post-infection community
data. We identified 6 OTUs (Enterobacteriaceae, Ruminococcaceae,
Lactobacillus, Bacteroides, Porphyromonadaceae, Erysipelotrichaceae)
that were important to the day 0 model and either the baseline or 1 day
post-infection models (Table S14). Thus, a subset of bacterial OTUs were
important for determining \emph{C. difficile} colonization dynamics
across multiple timepoints.

To determine whether the OTUs driving the classification models also
varied between sources, were altered by clindamycin treatment, or both,
we examined whether the top 20 ranked OTUs from each model overlapped
with OTUs that varied between sources (Fig. 1D, 3D, 4D and Tables S5,
S8, S10) or were impacted by clindamycin treatment (Fig. 3E and Table
S9). We found a subset of OTUs that were important to the baseline (day
-1), post-clindamycin (day 0), and 1 day post-infection models and
overlapped with bacteria that varied between sources, were altered by
clindamycin treatment, or both (Fig. S4). Combining the overall
comparison results for the 3 models identified 14 OTUs associated with
source, 21 OTUs associated with clindamycin treatment, and 6 OTUs
associated with both (Fig. 5B). Several OTUs (\emph{Bacteroides,
Enterococcus, Enterobacteriaceae, Porphyromonadaceae}) that overlapped
with our previous analyses appeared across at least 2 models, so we
examined how the relative abundances of these OTUs varied over the
course of the experiment (Fig. 6). Throughout the experiment, there was
at least 1 timepoint where relative abundances of these OTUs
significantly varied between sources (Table S15). Interestingly, there
were no OTUs that emerged as consistently enriched or depleted in mice
that were colonized past 7 days post-infection, suggesting multiple
bacteria influence \emph{C. difficile} colonization dynamics. Together,
these results suggest the initial bacterial communities and their
responses to clindamycin influence the clearance of \emph{C. difficile}.

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

By running our CDI model with mice from 6 different sources, we were
able to identify bacterial taxa that were unique to sources throughout
the experiment as well as taxa that were universally impacted by
clindamycin. We trained L2 logistic regression models with baseline (day
-1), post-clindamycin treatment (day 0), and 1-day post-infection fecal
community data that could predict whether mice cleared \emph{C.
difficile} by 7 days post-infection better than random chance. We
identified \emph{Bacteroides, Enterococcus, Enterobacteriaceae,
Porphyromonadaceae} (Fig. 6) as candidate bacteria within these
communities that were influencing variation in \emph{C. difficile}
colonization dynamics since these bacteria were all important in the
logistic regression models and varied by source, were impacted by
clindamycin treatment, or both. Overall, our results demonstrated
clindamycin was sufficient to render mice from multiple sources
susceptible to CDI and only a subset of the interindividual microbiota
variation across mice from different sources was associated with the
time needed to clear \emph{C. difficile}.

Differences in CDI mouse model studies have been attributed to
intestinal microbiota variation across sources. For example, groups
using the same clindamycin treatment and C57BL/6 mice had different
\emph{C. difficile} outcomes, one having sustained colonization (32),
while the other had transient (18), despite both using \emph{C.
difficile} VPI 10643. Baseline differences in the microbiota composition
have been hypothesized to partially explain the differences in
colonization outcomes and overall susceptibility to \emph{C. difficile}
after treatment with the same antibiotic (13, 31). The bacterial
perturbations induced by clindamycin treatment have been well
characterized and our findings agree with previous CDI mouse model work
demonstrating \emph{Enterococcus} and \emph{Enterobacteriaceae} were
associated with \emph{C. difficile} susceptibility and
\emph{Porpyhromonadaceae}, \emph{Lachnospiraceae},
\emph{Ruminococcaceae}, and \emph{Turicibacter} were associated with
resistance (19, 21, 32, 33, 43--46). While we have demonstrated that
susceptibility is uniform across sources of mice after clindamycin
treatment, there could be different outcomes for either susceptibility
or clearance in the case of other antibiotic treatments. The \emph{C.
difficile} strain used could also be contributing to the variation in
\emph{C. difficile} outcomes seen across different research groups. For
example, a group found differential colonization outcomes after
clindamycin treatment, with \emph{C. difficile} 630 and M68 infections
eventually becoming undectable while strain BI-7 remained detectable up
to 70 days post-treatment (45). We found the time needed to naturally
clear \emph{C. difficile} varied across sources of mice implying that at
least in the context of the same perturbation, microbiota differences
seemed to influence infection outcome more than susceptibility. More
importantly, we were able to reduce the variation observed across
sources to identify a subset of OTUs that were also important for
predicting \emph{C. difficile} colonization status 7 days
post-infection. Since all but 3 mice eventually cleared \emph{C.
difficile} 630 by 9 days post-infection and the model built with the
post-clindamycin (day 0) OTU relative abundance data had the best
performance, our results suggest clindamycin treatment had a larger role
in determining \emph{C. difficile} susceptibility and clearance than the
source of the mice.

Our approach successfully increased the diversity of murine bacterial
communities tested in our clindamycin \emph{C. difficile} model. One
alternative approach that has been used in some CDI studies (47--52) is
to associate mice with human microbiotas. However, a major caveat to
this method is the substantial loss of human microbiota community
members upon transfer to mice (53, 54). Additionally with the exception
of 2 recent studies (47, 48), most of the CDI mouse model studies to
date associated mice with just 1 types of human microbiota either from a
single donor or a single pool from multiple donors (49--52), which does
not aid in the goal of modeling the interpersonal variation seen in
humans to understand how the microbiota influences susceptiblity to CDIs
and adverse outcomes. Importantly, our study using mice from 6 different
sources increased the variation between groups of mice compared to using
1 source alone, to better reflect the inter-individual microbiota
variation observed in humans. Encouragingly, decreased
\emph{Bifidobacterium}, \emph{Porphyromonas}, \emph{Ruminococcaceae} and
\emph{Lachnospiraceae} and increased \emph{Enterobacteriaceae},
\emph{Enterococcus}, \emph{Lactobacillus}, and \emph{Proteus} have all
been associated with human CDIs (7) and were well represented in our
study, suggesting most of the mouse sources are suitable for gaining
insights into the bacteria influencing \emph{C. difficile} colonization
and infections in humans. An important exception was
\emph{Enterococcus}, which was primarily absent from University of
Michigan colonies and \emph{Proteus}, which was only found in Taconic
mice. Importantly, the fact that some CDI-associated bacteria were only
found in a subset of mice has important implications for future CDI
mouse model studies.

There are several limitations to our work. The microbiota is composed of
viruses, fungi, and parasites in addition to bacteria, and these
non-bacterial members can also vary across sources of mice (55, 56).
While our study focused solely on the bacterial portion, viruses and
fungi have also begun to be implicated in the context of CDIs or FMT
treatments for recurrent CDIs (35, 57--60). Beyond community
composition, the metabolic function of the microbiota also has a CDI
signature (20, 46, 61, 62) and can vary across mice from different
sources (63). For example, microbial metabolites, particularly secondary
bile acids and butyrate production, have been implicated as important
contributors to \emph{C. difficile} resistance (33, 45). Although, we
only looked at composition, \emph{Ruminococcaceae} and
\emph{Lachnospiraceae} both emerged as important taxa for classifying
day 7 \emph{C. difficile} colonization status and metagenomes from these
bacteria have been shown to contain the bile acid-inducible gene cluster
necessary for secondary bile acid formation and ability to produce
butyrate (50, 64). Interestingly, butyrate has previously been shown to
vary across mouse vendors and mediated resistance to \emph{Citrobacter
rodentium} infection, a model of enterohemorrhagic and enteropathogenic
\emph{Escherichia coli} infections (23). Evidence for immunological
toning differences in IgA and Th17 cells across mice from different
vendors have also been documented and (65, 66) could influence the host
response to CDI (67, 68). The outcome after \emph{C. difficile} exposure
depends on a multitude of factors, including age, diet, and immunity;
all of which are also influenced by the microbiota.

We have demonstrated that the ways baseline microbiotas from different
mouse sources respond to clindamycin treatment influences the length of
time mice remained colonized with \emph{C. difficile} 630. For those
interested in dissecting the contribution of the microbiota to \emph{C.
difficile} pathogenesis and treatments, using multiple sources of mice
may yield more insights than a single model alone. Furthermore, for
studies wanting to examine the interplay between a particular bactera
such as \emph{Enterococcus} and \emph{C. difficile}, these results could
serve as a resource for selecting which mice to order to address the
question. Using mice from multiple sources helps model the interpersonal
microbiota variation among humans to aid our understanding of how the
gut microbiota contributes to CDIs.

@@

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

This work was supported by the National Institutes of Health
(U01AI124255). ST was supported by the Michigan Institute for Clincial
and Health Research Postdoctoral Translation Scholars Program
(UL1TR002240). We thank members of the Schloss lab for feedback on
planning the experiments and data presentation, as well as code
tutorials and feedback through Code Club. In particular, we want to
thank Begüm Topçuoğlu for help with implementing L2 logistic regression
models using her
\href{https://github.com/SchlossLab/ML_pipeline_microbiome}{pipeline},
Ana Taylor for help with media preparation and sample collection, and
Nicholas Lesniak for his critical feedback on the manuscript. We also
thank members of Vincent Young's lab, particularly Kimberly Vendrov, for
guidance with the \emph{C. difficile} infection mouse model and donating
the mice. We also want to thank the Unit for Laboratory Animal Medicine
at the University of Michigan for maintaining our mouse colony and
providing the institutional support for our mouse experiments. Finally,
we thank Kwi Kim, Austin Campbell, and Kimberly Vendrov for their help
in maintaining the Schloss lab's anaerobic chamber.
